State of the Art in Concurrent Priority Queues \\ Student Paper for Seminar in Algorithms 2013W \\ Technical University of Vienna

### 2014-03-29 paper/paper.Rnw
Minor paper corrections

\begin{abstract}
Priority queues are abstract data structures which store a set of (priority, item) pairs
and allow efficient access to the item with minimal priority. Such queues are an important
element in various areas of computer science such as algorithmics (i.e. Dijkstra's shortest
path algorithm) and operating system (i.e. priority schedulers).

@@

There are several different criteria which allows reasoning about the correctness of concurrent
data structures. \emph{Linearizable} \cite{herlihy1990linearizability} operations appear to take
effect at a single instant in time at so-called linearization points. Linearizability is composable ---
any data structure composed of linearizable objects is also linearizable.
\emph{Sequential consistency} \cite{lamport1979make} requires the result of a set of operations
executed in parallel to be equivalent to the result of some sequential ordering of the same
operations.
\emph{Quiescently consistent} \cite{shavit1996diffracting} data structures
guarantee that the result of a set of parallel operations is equal to the result of a sequential ordering
after a period of quiescence, i.e. an interval without active operations has passed.

@@

\citeauthor{sundell2003fast} proposed the first lock-free concurrent priority queue in
\citeyear{sundell2003fast} \cite{sundell2003fast}. The data structure is linearizable
and is implemented using commonly available atomic primitives \ac{CAS}, \ac{TAS}, and \ac{FAA}.
In contrast to other structures covered in this paper, this priority queue is restricted to
contain items with pairwise distinct priorities. Inserting a new item with a priority already contained
in the list simply performs an update of the associated value.
A real-time version is also provided which will we will not discuss further (interested readers are
referred to \cite{sundell2003fast}).

@@

The structure of each node is basically identical to Figure \ref{fig:basicsl}. However, \citeauthor{sundell2003fast}
exploit the fact that the two least significant bits of pointers on 32- and 64-bit systems
are unused and reuse these as deletion marks. A set least significant bit signifies on a pointer signifies
that the current node is about to be deleted.
Reuse of \lstinline|node.level[i]| pointers
prevents situations in which a new node is inserted while its predecessor is being removed,
effectively deleting both nodes from the list. Likewise, the reuse of the \lstinline|node.value|
pointer ensures that updates of pointer values (which occur when a node with the inserted priority already exists)
handle concurrent node removals correctly.

@@

This implementation uses the lock-free memory management by invented by \citeauthor{valois1996lock}
\cite{valois1995lock,valois1996lock} and corrected by \citeauthor{michael1995correction} 
\cite{michael1995correction}. It was chosen in particular because this scheme can guarantee validity
of \lstinline|prev| as well as all \lstinline|next| pointers. Additionally, it does not require a separate
garbage collector thread.

@@

This mechanism requires a new invariant, in that the set of all logically deleted nodes must always
form a prefix of the SkipList. Recall that in the \citeauthor{sundell2003fast} queue, deletion flags
for node \lstinline|n| were packed into \lstinline|n.next| pointers, preventing insertion of new
nodes \emph{after} deleted nodes. This implementation instead places the deletion flag into the
lowest level \lstinline|next| pointer of the previous node, preventing inserations \emph{before}
logically deleted nodes.

@@

@@

\begin{itemize}
\item \textbf{GlobalLock} An instance of the \lstinline|std::priority_queue<T>| class provided
      by the C++ standard library, protected by a single global lock.
\item \textbf{Hunt} \citeauthor{hunt1996efficient} provide an implementation
      of their Heap-based design \cite{hunt1996efficient} at \url{ftp://ftp.cs.rochester.edu/pub/packages/concurrent_heap/}.
      However, as the original code was written for the MIPS architecture, we chose to
      use the alternative implementation in \emph{libcds}\footnote{\url{http://sourceforge.net/p/libcds/code/}}
      instead.
\item \textbf{Noble} An implementation of the \citeauthor{sundell2003fast} priority queue \cite{sundell2003fast}
      is available in the Noble library\footnote{\url{http://www.non-blocking.com/Eng/download-demo.aspx}}.
      \textbf{Noble} is lock-free, based on SkipLists, and limited to unique priorities.
      According to the authors, the commercial Noble library includes a more efficient
      version of this data structure that can also handle duplicate priority values.
\item \textbf{Linden} Code for the \citeauthor{linden2013skiplist} priority queue \cite{linden2013skiplist}
      is provided by the authors under an open source license\footnote{\url{http://user.it.uu.se/~jonli208/priorityqueue}}.
      It is lock-free, SkipList-based, and uses \citeauthor{fraser2004practical}'s lock-free
      SkipList implementation. The aim of this implementation is to minimize contention in
      calls to \lstinline|DeleteMin|.
\item \textbf{SprayList} A relaxed concurrent priority queue based on \citeauthor{fraser2004practical}'s
      SkipList using random walks to spread data accesses
      incurred by \lstinline|DeleteMin| calls. Code provided by \citeauthor{alistarhspraylist} is
      available on Github\footnote{\url{https://github.com/jkopinsky/SprayList}}.
\end{itemize}

@@

Recently invented relaxed priority queues do not exhibit the inherent bottleneck at the front of the list,
as they do not necessarily return the minimal element within the queue and are able to spread
\lstinline|DeleteMin| accesses over a larger area of the structure. In consequence, relaxed queues scale noticeable
better and to larger thread counts than strict designs. Further research is necessary in order to fully
explore the possibilities provided by relaxed data structures.

>>>>

\begin{abstract}
Priority queues are abstract data structures which store a set of key/value pairs
and allow efficient access to the item with the minimal (maximal) key. Such queues are an important
element in various areas of computer science such as algorithmics (i.e. Dijkstra's shortest
path algorithm) and operating system (i.e. priority schedulers).

@@

There are several different criteria which allow reasoning about the correctness of concurrent
data structures. \emph{Linearizable} \cite{herlihy1990linearizability} operations appear to take
effect at a single instant in time at so-called linearization points. Linearizability is composable ---
any data structure composed of linearizable objects is also linearizable.
\emph{Sequential consistency} \cite{lamport1979make} requires the result of a set of operations
executed in parallel to be equivalent to the result of some sequential ordering of the same
operations.
\emph{Quiescently consistent} \cite{shavit1996diffracting} data structures
guarantee that the result of a set of parallel operations is equal to the result of a sequential ordering
after a period of quiescence, i.e. an interval without active operations has passed.

@@

\citeauthor{sundell2003fast} proposed the first lock-free concurrent priority queue in
\citeyear{sundell2003fast} \cite{sundell2003fast}. The data structure is linearizable
and is implemented using commonly available atomic primitives \ac{CAS}, \ac{TAS}, and \ac{FAA}.
In contrast to other structures covered in this paper, this priority queue is restricted to
contain items with pairwise distinct priorities. Inserting a new item with a priority already contained
in the list simply performs an update of the associated value.
A real-time version is also provided which we will not discuss further (interested readers are
referred to \cite{sundell2003fast}).

@@

The structure of each node is basically identical to Figure \ref{fig:basicsl}. However, \citeauthor{sundell2003fast}
exploit the fact that the two least significant bits of pointers on 32- and 64-bit systems
are unused and reuse these as deletion marks. A set least significant bit on a pointer signifies
that the current node is about to be deleted.
Reuse of \lstinline|node.level[i]| pointers
prevents situations in which a new node is inserted while its predecessor is being removed,
effectively deleting both nodes from the list. Likewise, the reuse of the \lstinline|node.value|
pointer ensures that updates of pointer values (which occur when a node with the inserted priority already exists)
handle concurrent node removals correctly.

@@

This implementation uses the lock-free memory management invented by \citeauthor{valois1996lock}
\cite{valois1995lock,valois1996lock} and corrected by \citeauthor{michael1995correction} 
\cite{michael1995correction}. It was chosen in particular because this scheme can guarantee validity
of \lstinline|prev| as well as all \lstinline|next| pointers. Additionally, it does not require a separate
garbage collector thread.

@@

This mechanism requires a new invariant, in that the set of all logically deleted nodes must always
form a prefix of the SkipList. Recall that in the \citeauthor{sundell2003fast} queue, deletion flags
for node \lstinline|n| were packed into \lstinline|n.next| pointers, preventing insertion of new
nodes \emph{after} deleted nodes. This implementation instead places the deletion flag into the
lowest level \lstinline|next| pointer of the previous node, preventing insertions \emph{before}
logically deleted nodes.

@@

% TODO: Mention that the global list is periodically scanned by each thread.

@@

\begin{itemize}
\item \textbf{GlobalLock} An instance of the \lstinline|std::priority_queue<T>| class provided
      by the C++ standard library, protected by a single global lock.
\item \textbf{Hunt} \citeauthor{hunt1996efficient} provide an implementation
      of their Heap-based design \cite{hunt1996efficient} at \url{ftp://ftp.cs.rochester.edu/pub/packages/concurrent_heap/}.
      However, as the original code was written for the MIPS architecture, we chose to
      use the alternative implementation in \emph{libcds}\footnote{\url{http://sourceforge.net/p/libcds/code/}}
      instead.
\item \textbf{Noble} An implementation of the \citeauthor{sundell2003fast} priority queue \cite{sundell2003fast}
      is available in the Noble library\footnote{\url{http://www.non-blocking.com/Eng/download-demo.aspx}}.
      \textbf{Noble} is lock-free, based on SkipLists, and limited to unique priorities.
      According to the authors, the commercial Noble library includes a more efficient
      version of this data structure that can also handle duplicate priority values.
\item \textbf{Linden} Code for the \citeauthor{linden2013skiplist} priority queue \cite{linden2013skiplist}
      is provided by the authors under an open source license\footnote{\url{http://user.it.uu.se/~jonli208/priorityqueue}}.
      It is lock-free and uses \citeauthor{fraser2004practical}'s lock-free
      SkipList. The aim of this implementation is to minimize contention in
      calls to \lstinline|DeleteMin|.
\item \textbf{SprayList} A relaxed concurrent priority queue based on \citeauthor{fraser2004practical}'s
      SkipList using random walks to spread data accesses
      incurred by \lstinline|DeleteMin| calls. Code provided by \citeauthor{alistarhspraylist} is
      available on Github\footnote{\url{https://github.com/jkopinsky/SprayList}}.
\end{itemize}

@@

Recently invented relaxed priority queues do not exhibit the inherent bottleneck at the front of the list,
as they do not necessarily return the minimal element within the queue and are able to spread
\lstinline|DeleteMin| accesses over a larger area of the structure. In consequence, relaxed queues scale noticeably
better and to larger thread counts than strict designs. Further research is necessary in order to fully
explore the possibilities provided by relaxed data structures.
